MSCA 31006: Final Project - AQI Forecasting

### 2023-08-15 MSCA_31006_Time_Series_Final_Project - ARIMA.Rmd
Add files via upload

---
title: 'MSCA 31006: Final Project - AQI Forecasting'
author: "Aashish Singh, Jacob Brewer, Nikitha Gopal, Sydney Peters"
date: "8/10/2023"
output: html_document
---

@@

```{r, tidy=TRUE}
#path <- "/Users/aashishsingh/Desktop/Time Series - Final Project/"
#pm2_5_data <- read.csv(paste0(path, "la_pm2_5_pollutant_data.csv"), 
#                     na.strings=c("", "NA"))
#ozone_data <- read.csv(paste0(path, "la_ozone_pollutant_data.csv"), 
#                     na.strings=c("", "NA"))

path <- "D:/TS Final/"
pm2_5_data <- read.csv(paste0(path, "pm2.5_LA.csv"), 
                     na.strings=c("", "NA"))
ozone_data <- read.csv(paste0(path, "ozone_LA.csv"), 
                     na.strings=c("", "NA"))

@@

```
#### Step 5b: Extract values for time series and plot the series - Monthly

@@

>>>>

---
title: 'MSCA 31006: Final Project - AQI Forecasting'
author: "Aashish Singh, Jacob Brewer, Nikitha Gopal, Sydney Peters"
date: "8/10/2023"
output: html_document
---
```{r}
# List all objects in the environment
objects_to_remove <- ls()

# Remove all objects
rm(list = objects_to_remove)
```

@@

```{r, tidy=TRUE}
path <- "/Users/nikithagopal/Desktop/MSCA Data/Time Series Final/"
pm2_5_data <- read.csv(paste0(path, "la_pm2_5_pollutant_data.csv"), 
                     na.strings=c("", "NA"))
ozone_data <- read.csv(paste0(path, "la_ozone_pollutant_data.csv"), 
                     na.strings=c("", "NA"))

#path <- "/Users/sydneypeters/Desktop/"
#pm2_5_data <- read.csv(paste0(path, "pm2.5_LA.csv"), 
#                     na.strings=c("", "NA"))
#ozone_data <- read.csv(paste0(path, "ozone_LA.csv"), 
#                     na.strings=c("", "NA"))

@@

#### Step 5b: Extract values for time series and plot the series - Monthly

@@

#### Step 10a: ARIMA model using auto.arima() - Weekly
```{r}
eacf <-
function (z,ar.max=7,ma.max=13) 
{
#
#  PROGRAMMED BY K.S. CHAN, DEPARTMENT OF STATISTICS AND ACTUARIAL SCIENCE,
#  UNIVERSITY OF IOWA.
#
#  DATE: 4/2001
#  Compute the extended sample acf (ESACF) for the time series stored in z.
#  The matrix of ESACF with the AR order up to ar.max and the MA order
#  up to ma.max is stored in the matrix EACFM.
#  The default values for NAR and NMA are 7 and 13 respectively.
#  Side effect of the eacf function:
#  The function prints a coded ESACF table with
#  significant values denoted by * and nosignificant values by 0, significance
#  level being 5%.
#
#  Output:
#   eacf=matrix of esacf
#   symbol=matrix of coded esacf
#

lag1<-function(z,lag=1){c(rep(NA,lag),z[1:(length(z)-lag)])}
reupm<-function(m1,nrow,ncol){
k<-ncol-1
m2<-NULL
for (i in 1:k){
i1<-i+1
work<-lag1(m1[,i])
work[1]<--1
temp<-m1[,i1]-work*m1[i1,i1]/m1[i,i]
temp[i1]<-0
m2<-cbind(m2,temp)
}
m2}
ceascf<-function(m,cov1,nar,ncol,count,ncov,z,zm){
result<-0*seq(1,nar+1)
result[1]<-cov1[ncov+count]
for (i in 1:nar) {
temp<-cbind(z[-(1:i)],zm[-(1:i),1:i])%*%c(1,-m[1:i,i])
result[i+1]<-acf(temp,plot=FALSE,lag.max=count,drop.lag.0=FALSE)$acf[count+1]
}
result
}

ar.max<-ar.max+1
ma.max<-ma.max+1
nar<-ar.max-1
nma<-ma.max
ncov<-nar+nma+2
nrow<-nar+nma+1
ncol<-nrow-1
z<-z-mean(z)
zm<-NULL
for(i in 1:nar) zm<-cbind(zm,lag1(z,lag=i))
cov1<-acf(z,lag.max=ncov,plot=FALSE,drop.lag.0=FALSE)$acf
cov1<-c(rev(cov1[-1]),cov1)
ncov<-ncov+1
m1<-matrix(0,ncol=ncol,nrow=nrow)
for(i in 1:ncol) m1[1:i,i]<-
ar.ols(z,order.max=i,aic=FALSE,demean=FALSE,intercept=FALSE)$ar
eacfm<-NULL
for (i in 1:nma) {
m2<-reupm(m1=m1,nrow=nrow,ncol=ncol)
ncol<-ncol-1
eacfm<-cbind(eacfm, ceascf(m2,cov1,nar,ncol,i,ncov,z,zm))
m1<-m2}
work<-1:(nar+1)
work<-length(z)-work+1
symbol<-NULL
for ( i in 1:nma) {
work<-work-1
symbol<-cbind(symbol,ifelse(abs(eacfm[,i])>2/work^.5, 'x','o'))}
rownames(symbol)<-0:(ar.max-1)
colnames(symbol)<-0:(ma.max-1)
cat('AR/MA\n')
print(symbol,quote=FALSE)
invisible(list(eacf=eacfm,ar.max=ar.max,ma.ma=ma.max,symbol=symbol))
}
```

```{r}
eacf(train_weekly)
# p <- 1,2,3
# q <- 2,3,4
```

```{r}

kpss_test_result <- kpss.test(train_weekly_diff)
adf_test_result <- adf.test(train_weekly_diff)

cat("KPSS: ", round(kpss_test_result$p.value, 2), "\n")
cat("ADF: ", round(adf_test_result$p.value, 2), "\n")

# Fit ARIMA model using auto.arima()
# Find appropriate values for d and D using ndiffs and nsdiffs
# d <- ndiffs(train_weekly_diff)
# D <- nsdiffs(train_weekly_diff)

# Fit ARIMA model using auto.arima() with seasonal components and previously found d and D
arima_model_weekly <- auto.arima(train_weekly, seasonal = TRUE, stepwise = FALSE,
                                 approximation = FALSE, trace = FALSE, lambda = "auto",
                                 start.p = 1, max.p = 4,
                                 start.q = 2, max.q = 5, 
                                 d=0)

# auto.arima(pm2_5_weekly_diff, lambda = "auto", seasonal=TRUE, )

# Forecast using ARIMA model
h <- length(test_weekly)
forecast_arima_weekly <- forecast(arima_model_weekly, h = h)

plot(forecast_arima_weekly, main = "PM 2.5 AQI Forecast - sARIMA",
         xlab = "Week", ylab = "PM 2.5 AQI")
lines(test_weekly)

```
```{r}
forecast_arima_weekly
```

```{r}

# Check the residuals of the ARIMA model
checkresiduals(arima_model_weekly)

# Calculate MAE
mae <- mean(abs(forecast_arima_weekly$mean - test_weekly))
# Calculate MAPE
mape <- mean(abs((forecast_arima_weekly$mean - test_weekly) / test_weekly)) * 100

# Get AICc and BIC values from the ARIMA model
aic <- arima_model_weekly$aic
bic <- arima_model_weekly$bic

# Corrected AICc calculation
n <- length(train_weekly)
k <- length(arima_model_weekly$coef)
aic_corrected <- aic + 2 * (k + 1) * (k + 2) / (n - k - 2)

# Print the results
cat("MAE: ", round(mae, 2), "\n")
cat("MAPE: ", round(mape, 2), "%\n")
cat("AICc: ", round(aic_corrected, 2), "\n")
cat("BIC: ", round(bic, 2), "\n")

# Create a time index for the forecasted values
forecast_time_index <- seq(from = end(test_weekly) + 1 / 52, 
                           length.out = length(forecasted_values), by = 1 / 52)

# Plot historical data, actual test, and forecast
plot(train_weekly, xlim=c(1999, 2023), ylim=c(min(train_weekly), max(train_weekly)), main="Train, Test, and Forecasted Data - Weekly", xlab="Year", ylab="PM2.5 AQI Value")
lines(test_weekly, col="red") # Test data in red
lines(forecast_arima_weekly$mean, col="blue", type="o") # Forecasted data in blue
legend("topleft", legend=c("Train", "Test", "Forecast"), fill=c("black", "red", "blue"))

comparison_df <- data.frame(
  Date = time(test_weekly),
  Actual = as.vector(test_weekly),
  Forecasted = as.vector(forecast_arima_weekly$mean)
)
print(comparison_df)
```
```{r}
summary(arima_model_weekly)
```

```{r}
# Assuming your residuals are stored in a variable called "residuals"
shapiro_test_result <- shapiro.test(arima_model_weekly$residuals)

# Print the test result
print(shapiro_test_result)

# Check the p-value
if (shapiro_test_result$p.value < 0.05) {
  cat("The residuals are not normally distributed (p-value:", shapiro_test_result$p.value, ")\n")
} else {
  cat("The residuals appear to be normally distributed (p-value:", shapiro_test_result$p.value, ")\n")
}
```

#### ARIMA model using auto.arima() - Monthly

```{r}
# Convert monthly data to a time series object
# pm2_5_monthly_ts <- ts(pm2_5_monthly["19990131/20230811"], start = c(1999, 1), frequency = 12)

y_values_monthly <- train_monthly
optimal_lambda <- BoxCox.lambda(y_values_monthly)

# Apply the Box-Cox transformation
transformed_data_monthly <- if (optimal_lambda != 0) {
  (y_values_monthly^optimal_lambda - 1) / optimal_lambda
} else {
  log(y_values_monthly)
}

# Check stationarity, perform differencing if needed
kpss_test_result <- kpss.test(pm2_5_monthly_ts)
adf_test_result <- adf.test(pm2_5_monthly_ts)

if (kpss_test_result$p.value < 0.05 || adf_test_result$p.value < 0.05) {
  pm2_5_monthly_diff <- diff(pm2_5_monthly_ts)
} else {
  pm2_5_monthly_diff <- pm2_5_monthly_ts
}

# Check stationarity, perform differencing if needed
kpss_test_result <- kpss.test(transformed_data_monthly)
adf_test_result <- adf.test(transformed_data_monthly)

if (kpss_test_result$p.value < 0.05 || adf_test_result$p.value < 0.05) {
  pm2_5_monthly_diff <- diff(transformed_data_monthly)

  # Check stationarity after first differencing
  kpss_test_result_diff <- kpss.test(pm2_5_monthly_diff)
  adf_test_result_diff <- adf.test(pm2_5_monthly_diff)

  if (kpss_test_result_diff$p.value < 0.05 || adf_test_result_diff$p.value < 0.05) {
    # If still non-stationary, apply second-order differencing
    pm2_5_monthly_diff <- diff(pm2_5_monthly_diff)

    print("First and Second Order Differencing was performed")
  }
} else {
  pm2_5_monthly_diff <- transformed_data_monthly
  print("First Order Differencing was performed")
}

# Fit ARIMA model using auto.arima()
arima_model_monthly <- auto.arima(pm2_5_monthly_diff, seasonal = TRUE)

# Forecast using ARIMA model
h <- length(test_monthly)
forecast_arima_monthly <- forecast(arima_model_monthly, h = h)

original_scale_forecasts_monthly <- InvBoxCox(forecast_arima_monthly$mean, lambda = optimal_lambda)

print(cat("Optimal Lambda Value:", optimal_lambda, "\n"))

# Plot the original data
original_scale_forecasts_monthly <- original_scale_forecasts_monthly * 100
plot(train_monthly, main = "Original Data and Forecasts")
lines(original_scale_forecasts_monthly, col = "blue")  # Add forecast line in blue

# Add legend
legend("topleft", legend = c("Original Data", "Forecasts"), col = c("black", "blue"), lwd = 1)

```
```{r}

checkresiduals(arima_model_monthly)
#Ljung-Box test .05179

plot(pm2_5_monthly_diff, xlim=c(1999, 2023), ylim=c(min(pm2_5_monthly_diff), max(pm2_5_monthly_diff)), main="Train, Test, and Forecasted Data - Monthly", xlab="Year", ylab="PM2.5 AQI Value")
lines(test_monthly, col="red") # Test data in red
lines(original_scale_forecasts_monthly, col="blue", type="o") # Forecasted data in blue
legend("topleft", legend=c("Train", "Test", "Forecast"), fill=c("black", "red", "blue"))

mae <- mean(abs(test_monthly - original_scale_forecasts_monthly))
print(cat("Mean Absolute Error (MAE):", round(mae, 2), "\n"))
#Mean Absolute Error (MAE): 9.31

comparison_df <- data.frame(
  Date = time(test_monthly),
  Actual = as.vector(test_monthly),
  Forecasted = as.vector(original_scale_forecasts_monthly)
)
print(comparison_df)
